---
title: "Project Proposal"
author: "Dan Fanelli"
date: "March 7, 2016"
output: html_document
---

```{r include=FALSE, cache=FALSE}
# DO NOT REMOVE
# THIS IS FOR SETTING SOME PLOTTING PARAMETERS SO THAT YOUR PLOTS DON'T TAKE UP TOO MUCH SPACE
# IF YOU WOULD LIKE TO CHANGE THESE, SEE HELP FILES ON THE par() FUNCTION
# OR ASK FOR HELP
library(knitr)
## set global chunk options
opts_chunk$set(fig.path='figure/manual-', cache.path='cache/manual-', fig.align='center', fig.show='hold', par=TRUE)
## tune details of base graphics (http://yihui.name/knitr/hooks)
knit_hooks$set(par=function(before, options, envir){
if (before && options$fig.show!='none') par(mar=c(4,4,.2,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
})
```

```{r}
# load data
# data_url <- "ftp://ftp.fec.gov/FEC/Presidential_Map/2016/P00000001/P00000001-ALL.zip"

data_url <- "data_project_subset.csv"
```


### Research question 

*(You should phrase your research question in a way that matches up with the scope of inference your dataset allows for)*

**What can we learn about our presidential election process by the characteristics of the presidential campaign contribution records?**

### Cases 

**What are the cases, and how many are there?**

* There are about 1.2 million cases (and growing I believe...)
* Each case represents the individual contribution transaction

### Data collection 

**Describe the method of data collection.**

* CSV download

### Type of study 

**What type of study is this (observational/experiment)?**

* This is an observational study.

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

* Download "ALL.zip" from http://www.fec.gov/disclosurep/PDownload.do

### Response 

**What is the response variable, and what type is it (numerical/categorical)?**

* The contribution amount will be the response variable. 

### Explanatory 

**What is the explanatory variable, and what type is it (numerical/categorival)?**

* The explanatory variables may include CANDIDATE NAME, CONTRIBUTOR OCCUPATION, CONTRIBUTOR EMPLOYER, CONTRIBUTOR CITY, and CONTRIBUTOR STATE.
* These explanatory variables are categorical

### Relevant summary statistics 

**Provide summary statistics relevant to your research question. For example, if you are comparing means across groups provide means, SDs, sample sizes of each group. This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r message=FALSE, warning=FALSE}
library(sqldf)

campaign_data <- read.csv(data_url, header = TRUE)

sqldf("select cand_nm, contbr_employer, sum(CONTB_RECEIPT_AMT) as sum, count(*) from campaign_data group by cand_nm, contbr_employer order by sum desc, count(*) desc LIMIT 20")

```

